---
title: "Comparing epidemic scenarios"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Comparing epidemic scenarios}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 150
)
```

```{r setup}
library(scenarios)
library(ggplot2)
```

```{r}
# UK population size is about 67 million in 2023
uk_pop <- 67 * 1e6

# R0 for pandemic influenza
r0_flu <- 1.5
```

This first example considers two scenarios of the final size of an influenza epidemic, and examines how considering the age structure of the population can influence estimates of the final epidemic size.
  1. Scenario 1 is the uniform mixing scenario, in which all individuals are assumed to be mixing homogeneously, with no differences in social contacts by age or other demographic characteristics.
  2. Scenario 2 is the age-stratified mixing scenario, in which individuals are divided into three age bins, 0 -- 19, 20 -- 39, and 40+, with different social contact patterns for each age group with the other age groups.

```{r}
# create a scenario object for uniform mixing and susceptibility
uniform_mixing <- scenario(
  name = "uniform_mixing",
  model_function = "finalsize::final_size",
  parameters = list(
    r0 = r0_flu,
    contact_matrix = matrix(data = 1.0 / uk_pop),
    demography_vector = uk_pop,
    susceptibility = matrix(1.0),
    p_susceptibility = matrix(1.0),
    solver = "newton"
  ),
  extra_info = list(
    country = "UK",
    age_groups = "uniform population"
  ),
  replicates = 1
)

# a scenario for age stratified mixing
# get age group names
age_groups <- rownames(make_parameters_finalsize_UK()$contact_matrix)

age_stratified_mixing <- scenario(
  name = "age_stratified_mixing",
  model_function = "finalsize::final_size",
  parameters = make_parameters_finalsize_UK(r0 = r0_flu), # use convenience fn
  extra_info = list(
    country = "UK",
    age_groups = age_groups
  ),
  replicates = 1
)

# view scenario
age_stratified_mixing
```

```{r}
# create a third scenario from a different country, say, Germany
# with uniform mixing and susceptibility
# create a scenario object for uniform mixing
de_pop <- 84 * 1e6 # about 84 million
uniform_mixing_de <- scenario(
  name = "uniform_mixing_de",
  model_function = "finalsize::final_size",
  parameters = list(
    r0 = r0_flu,
    contact_matrix = matrix(data = 1.0 / de_pop),
    demography_vector = de_pop,
    susceptibility = matrix(1.0),
    p_susceptibility = matrix(1.0),
    solver = "newton"
  ),
  extra_info = list(
    age_groups = "uniform population"
  ),
  replicates = 1
)

# print to view
uniform_mixing_de

# add extra information to the Germany scenario using a helper function
uniform_mixing_de <- sce_add_info(
  uniform_mixing_de,
  info = list(
    country = "Germany"
  )
)

# check that information has been added
uniform_mixing_de
```

Create a comparison object to hold the comparisons and filter for comparable scenarios.

```{r}
flu_comparison <- comparison(
  uniform_mixing, age_stratified_mixing, uniform_mixing_de,
  baseline = "uniform_mixing"
)

# print to view
flu_comparison
```

Prepare data for all scenarios at once using the `run_scenario()` method for `comparison` objects.

```{r}
flu_comparison <- run_scenario(flu_comparison)

# check that data is ready
flu_comparison
```

## Comparing within the UK

To compare within the UK, filter the `comparison` object by the "country" entry in the extra information in each scenario.

```{r}
# filter on the country and whether data have a 'p_infected' column
flu_comparison_uk <- sce_filter_comparable(
  flu_comparison,
  match_variables = "country",
  comparison_variables = "p_infected"
)

# print and view
flu_comparison_uk

sce_get_scenario_names(flu_comparison_uk)
```

Get scenario outcomes and visualise.

```{r}
# get data which are final_size outputs bound together
data_comparison_uk <- sce_get_outcomes(flu_comparison_uk)

# print to view
data_comparison_uk

# change name of total population group
data_comparison_uk[scenario_name == "uniform_mixing", demo_grp := "Total pop."]
```

```{r}
ggplot(data_comparison_uk) +
  geom_col(
    aes(demo_grp, p_infected, fill = scenario_name),
    col = "black"
  ) +
  expand_limits(
    x = c(0.5, nrow(data_comparison_uk) + 0.5)
  ) +
  scale_fill_brewer(
    palette = "Accent",
    name = "Population susceptibility",
    labels = c("Heterogeneous", "Uniform")
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.key.height = unit(2, "mm"),
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  labs(
    x = "Age group",
    y = "% Infected"
  )
```

## Comparing uniform mixing, across countries

Filter for the "age_groups" attribute to compare uniform_mixing across the UK and Germany.

```{r}
# filter on the country and whether data have a 'p_infected' column
flu_comparison_uniform <- sce_filter_comparable(
  flu_comparison,
  match_variables = "age_groups",
  comparison_variables = "p_infected"
)

# print and view
flu_comparison_uniform

# print scenario names in this comparison
sce_get_scenario_names(flu_comparison_uniform)
```

Get scenario outcomes and visualise.

```{r}
# get data which are final_size outputs bound together
data_comparison_uniform <- sce_get_outcomes(flu_comparison_uniform)

# print to view
data_comparison_uniform
```

```{r}
ggplot(data_comparison_uniform) +
  geom_col(
    aes(demo_grp, p_infected, fill = scenario_name),
    col = "black",
    position = position_dodge(width = 1)
  ) +
  expand_limits(
    x = c(0.5, 1.5)
  ) +
  scale_fill_brewer(
    palette = "Dark2",
    name = "Country",
    labels = c("UK", "DE")
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.key.height = unit(2, "mm"),
    axis.text.x = element_blank()
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  labs(
    x = "Total population",
    y = "% Infected"
  )
```
