---
title: "Defining epidemic scenarios and comparisons"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Defining epidemic scenarios and comparisons}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This is a bare-bones vignette showing the basic functionality of _scenarios_. This package is a work in progress, and more functionality and documentation, including updates to this vignette, will be added in the near future.

::: {.alert .alert-warning}
**New to epidemic modelling in R?** It may help to read [The Epidemiologist R Handbook](https://epirhandbook.com/en/epidemic-modeling.html) for a primer on scenario modelling.
:::

::: {.alert .alert-primary}
## Use case {-}

To **compare** the outcomes of epidemic scenario models, such as the proportion of the population infected at a stage of the outbreak. Current use cases focus on comparing the cumulative proportion of individuals infected over the course of an epidemic wave, called the 'final size' of the epidemic.
:::

::: {.alert .alert-secondary}
### What you need {-}

  1. An epidemic model R function that produces data frames as outputs,
  2. Parameters to the epidemic model, which are typically arguments to the model function.

### What we assume {-}

  1. That the epidemic scenarios are broadly comparable on a conceptual level.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scenarios)
```

## Components of an epidemic scenario model

The _scenarios_ package is intended to help with the comparison of epidemic model scenarios.
The _scenarios_ package defines the `scenario` class, which is an epidemic model definition with or without its associated data.
Each `scenario` is expected to have:

1. A model function, 
2. The model parameters in the form of named function arguments, and
3. The number of replicates to run.

<!-- ::: {.alert .alert-secondary}
### Example of epidemic scenario model

::: -->

## Defining scenarios for the final size of an epidemic

<!-- Text about finalsize calculations here, with a box - link to finalsize website.
Set the scene for when this sort of comparison is likely to be necessary --- new epidemic outbreak, quick comparison of potential outcomes wanted. -->

This example considers the definition of scenario specifications for the outbreak of an epidemic in the U.K. population, and uses demography and social contact data from the POLYMOD survey [@mossong2008].
These are conveniently provided in the _finalsize_ package, and can be obtained for other coutries (and other surveys) from the _[socialmixr](https://epiforecasts.io/socialmixr/)_ package.

First define epidemic scenarios.

```{r}
# get inbuilt parameters using a convenience function
parameters_finalsize_low_r0 <- make_parameters_finalsize_UK(r0 = 1.5)
age_groups <- rownames(parameters_finalsize_low_r0$contact_matrix)

# define a low R0 scenario comparable to pandemic influenza
low_r0 <- scenario(
  model_function = "finalsize::final_size",
  parameters = parameters_finalsize_low_r0,
  extra_info = list(
    age_groups = age_groups
  ),
  replicates = 1L
)

# define a higher R0 scenario comparable to Covid-19
# note no age group information is added yet
high_r0 <- scenario(
  model_function = "finalsize::final_size",
  parameters = make_parameters_finalsize_UK(r0 = 5.0),
  replicates = 1L
)
```

```{r}
# print scenarios to examine information
low_r0
high_r0
```

## Getting scenario parameters

Scenario objects can be queried for their information using `sce_get_information()`. This function searches the scenario's `parameters` and `extra_info` lists for named elements and returns them if found.

```{r}
# get information on the model parameter r0
sce_get_information(low_r0, which = "r0")

# get information on the demography groups
sce_get_information(low_r0, which = c("r0", "age_groups"))

sce_get_information(low_r0, which = "contact_matrix")
```

When the scenario does not have this information, an informative error message is returned.

```{r error=TRUE}
# no data found
sce_get_information(high_r0, which = "age_groups")
```

## Adding information to scenarios

Information can be added to scenarios using `sce_add_info()`, with the information passed as an element of a named list. This function checks whether the list names are already present in the `extra_info` list, and adds them if they are not present, erroring otherwise.

```{r}
# add age group size information
high_r0 <- sce_add_info(
  high_r0, list(age_groups = age_groups)
)

# print to check that data are added
high_r0
```

## Prepare scenario data

The scenario objects created thus far do not have any data, and this is because the `model_function` has not been called with the `parameters` to populate the `data` list.

The scenario outcome data can be added to the scenario by running the scenario using `run_scenario()`.

```{r}
# add outcome data to scenarios
# assigning to different object for clarity
low_r0_data <- run_scenario(low_r0)
high_r0_data <- run_scenario(high_r0)
```

```{r}
# print scenario to check information
low_r0_data
```

## Getting scenario data

The scenario data can be accessed (so long as they are data frames) using `sce_get_outcomes()`, which returns a `data.table` of outcomes, with an additional column to indicate the replicate number.

```{r}
# peeking at data to examine column names
sce_peek_outcomes(low_r0_data, view_rows = FALSE)

# get full outcome data as data.table
sce_get_outcomes(low_r0_data)
```

## Check whether scenarios are comparable

```{r}
# comparing scenarios while expecting an identical match
# this is less useful when comparing scenarios that obviously
# differ on an important parameter such as R0
sce_are_comparable(
  baseline = low_r0_data,
  compare = high_r0_data,
  match_variables = "r0",
  comparison_variables = "p_infected"
)

# match variables must be *identical*
sce_are_comparable(
  baseline = low_r0_data,
  compare = high_r0_data,
  match_variables = "age_groups",
  comparison_variables = "p_infected"
)
```

## Creating a comparison object

Scenarios can be wrapped into a `comparison` object. This is a convenient and structured way of holding numerous scenarios. A brief introduction is given here.

```{r}
# create object, passing prepared scenarios with some names
outbreak_comparison <- comparison(
  pandemic_flu = low_r0_data, covid19 = high_r0_data,
  baseline = "pandemic_flu"
)

# check the scenario names in this comparison
sce_get_scenario_names(outbreak_comparison)

# print to examine output
outbreak_comparison
```

The data from individual scenarios can be accessed as a single data table using the `sce_get_outcomes()` function.

```{r}
# scenario output can be accessed from a comparison object as well
# using multiple dispatch on the sce_get_outcomes function
sce_get_outcomes(outbreak_comparison)
```

## References
